---
title: "STAT 1361 Final Project"
author: "Chris Park (chp117@pitt.edu - 4288421)"
date: "4/19/2022"
output:
  word_document: default
  pdf_document:
    df_print: paged
---

# Packages

```{r warning=FALSE, include=FALSE}
library(stats)
library(ggplot2)
library(glmnet)
library(pls)
library(leaps)
library(class)
library(boot)
library(splines)
library(gam)
library(splines)
library(tree)
library(randomForest)
library(BART)
library(gbm)
```

# Data

```{r}
# Reading in data set
data = read.table("train.csv", header = T, sep = ",")
```

# Exploring Dataset

```{r}
# Checking to see if there are missing data
sum(is.na(data))
# Checking to see if each instance of data is unique
nrow(unique(data))
# Checking data types of columns
str(data)
# Converting data types of certain columns
data$Date = as.Date(data$Date, "%d/%m/%Y")
data$Seasons = as.factor(data$Seasons)
data$Holiday = as.factor(data$Holiday)
data$Functioning = as.factor(data$Functioning)
```

```{r}
# Numerical Summary
summary(data)
# Correlations of numerical columns
numeric = sapply(data, is.numeric)
cor(data[,numeric])
```

```{r}
# Scatter plots showing the quantitative predictors' relationship to Count
par(mfrow = c(2,2))
plot(data$Date, data$Count)
plot(data$Hour, data$Count)
plot(data$Temperature, data$Count)
plot(data$Humidity, data$Count)
plot(data$Wind, data$Count)
plot(data$Visibility, data$Count)
plot(data$Dew, data$Count)
plot(data$Solar, data$Count)
plot(data$Rainfall, data$Count)
plot(data$Snowfall, data$Count)
# Box plots showing the qualitative predictors' relationship to Count
par(mfrow = c(1,3))
boxplot(Count ~ Seasons, data = data, las = 2)
boxplot(Count ~ Holiday, data = data)
boxplot(Count ~ Functioning, data = data)
```

```{r}
# Histograms
par(mfrow=c(2,2))
hist(data$Count, breaks = 20)
hist(data$Date, breaks = 20)
hist(data$Temperature, breaks = 20)
hist(data$Humidity, breaks = 20)
hist(data$Wind, breaks = 20)
hist(data$Visibility, breaks = 20)
hist(data$Dew, breaks = 20)
hist(data$Solar, breaks = 20)
hist(data$Rainfall, breaks = 20)
hist(data$Snowfall, breaks = 20)
```

# Splitting Untransformed Data into Training and Test Sets

```{r}
set.seed(1)
# 70-30 Split
train = sample(1:nrow(data), nrow(data)*.7)
df.train = data[train,]
df.test = data[-train,]
```

# Multiple Linear Regression Model (no changes)

```{r}
# Model
lm.mod = lm(Count ~ . -ID, data = df.train)
summary(lm.mod)
par(mfrow = c(2,2))
plot(lm.mod)
# MSE
lm.pred = predict(lm.mod, newdata = df.test)
MSE = mean((df.test$Count - lm.pred)^2)
MSE
```

# Non-linear Relationships

```{r}
# Polynomials for Numerical Values
lm.fit = lm(Count ~ Date + poly(Hour, 5) + poly(Temperature, 5) + poly(Humidity, 5) + poly(Wind, 5) + poly(Visibility, 5) + poly(Dew, 5) + poly(Solar, 5) + poly(Rainfall, 5) + poly(Snowfall, 5) + Seasons + Holiday + Functioning, data = df.train)
summary(lm.fit)
```

# ANOVA

```{r}
# Normal Model excluding ID
anova1 = gam(Count ~ . -ID, data = df.train)
# Model with Polynomials
anova2 = gam(Count ~ Date + s(Hour, 5) + s(Temperature, 5) + s(Humidity, 5) + s(Wind, 5) + s(Visibility, 5) + s(Dew, 5) + s(Solar, 5) + s(Rainfall, 5) + s(Snowfall, 5) + Seasons + Holiday + Functioning, data = df.train)
anova(anova1, anova2, test = "F")
summary(anova2)
```

# Transforming Predictor Variables

```{r}
# Copying Data
transformed.data = data
set.seed(1)
# 70-30 Split
train2 = sample(1:nrow(transformed.data), nrow(transformed.data)*.7)
df.train = data[train2,]
df.test = data[-train2,]

# Training
# Count: Right Skewed
df.train$Count = sqrt(df.train$Count)
# Wind: Right Skewed
df.train$Wind = sqrt(df.train$Wind)
# Visibility: Left Skewed
df.train$Visibility = (df.train$Visibility)^2
# Solar: Right Skewed
df.train$Solar = sqrt(df.train$Solar)
# Rainfall: Right Skewed
df.train$Rainfall = sqrt(df.train$Rainfall)
# Snowfall: Right Skewed
df.train$Snowfall = sqrt(df.train$Snowfall)

# Test
# Count: Right Skewed
df.test$Count = sqrt(df.test$Count)
# Wind: Right Skewed
df.test$Wind = sqrt(df.test$Wind)
# Visibility: Left Skewed
df.test$Visibility = (df.test$Visibility)^2
# Solar: Right Skewed
df.test$Solar = sqrt(df.test$Solar)
# Rainfall: Right Skewed
df.test$Rainfall = sqrt(df.test$Rainfall)
# Snowfall: Right Skewed
df.test$Snowfall = sqrt(df.test$Snowfall)

# Histogram of Transformed Variables
par(mfrow = c(3,2))
hist(df.train$Count)
hist(df.train$Wind)
hist(df.train$Visibility)
hist(df.train$Solar)
hist(df.train$Rainfall)
hist(df.train$Snowfall)
```

# Multiple Linear Regression Model on Transformed Data

```{r}
# Model
lm.mod = lm(Count ~ . -ID, data = df.train)
summary(lm.mod)
# MSE
lm.pred = predict(lm.mod, newdata = df.test)
MSE.lm0 = mean((df.test$Count - lm.pred)^2)
MSE.lm0
```

# Permutation Test for Insignificant Predictors from Model

```{r}
set.seed(1)
# MSE from Model
MSE.0 = mean((df.test$Count - lm.pred)^2)
# Number of permutations
nperm = 100
# Vector to hold permutation values
f.perm = rep(0,nperm)
# Loop to get MSE values of the permutations
for (i in 1:nperm) {
  Count = df.train$Count
  Date = df.train$Date
  Hour = df.train$Hour
  Temperature = df.train$Temperature
  Humidity = df.train$Humidity
  Wind = df.train$Wind
  Visibility = sample(df.train$Visibility)
  Dew = sample(df.train$Dew)
  Solar = sample(df.train$Solar)
  Rainfall = df.train$Rainfall
  Snowfall = sample(df.train$Snowfall)
  Seasons = df.train$Seasons
  Holiday = df.train$Holiday
  Functioning = df.train$Functioning
  perm.data = data.frame(Count, Date, Hour, Temperature, Humidity, Wind, Visibility, Dew, Solar, Rainfall, Snowfall, Seasons, Holiday, Functioning)
  model = lm(Count ~ ., data = perm.data)
  f.perm[i] = mean((df.test$Count - predict(model, df.test))^2)
}
# Histogram of vector
hist(f.perm)
# Line to indicate original MSE value
abline(v=MSE.0, col='blue', lwd=2)
# Calculating p-value
p = mean(f.perm < MSE.0)
p
```

# Interaction Terms

```{r}
# Dew Visibility, Solar Snowfall, Solar Visibility, Dew Solar
lm.mod2 = lm(Count ~ . -ID + Dew*Visibility + Solar*Snowfall + Solar*Visibility + Dew*Solar, data = df.train)
summary(lm.mod2)
# MSE
lm.pred2 = predict(lm.mod2, newdata = df.test)
MSE.lm.int = mean((df.test$Count - lm.pred2)^2)
MSE.lm.int
```

# Best Subsets Selection on Training Set

```{r}
set.seed(1)
# Number of predictors
p = 15
# Model Selection
regfit.full = regsubsets(Count ~ . -ID, data = df.train, nvmax = p)
reg.summary = summary(regfit.full)
# names(reg.summary)
# Model Matrix
train.mat = model.matrix(Count ~ . -ID, data = df.train, nvmax = p)
# Vector to store training MSE
val.errors = rep(NA, p)
# Looping through number of predictors
for (i in 1:p) {
  # Coefficients of models with i predictors
  coefi = coef(regfit.full, id = i)
  pred = train.mat[,names(coefi)] %*% coefi
  val.errors[i] = mean((pred - df.train$Count)^2)
}
# Plotting train MSE
plot(val.errors, xlab = "# of predictors", ylab = "training MSE")

par(mfrow=c(2, 2))
# Adj R^2
plot(reg.summary$adjr2, xlab = "Number of Variables", ylab = "Adjusted RSq", type = "l")
max.adjr2 = which.max(reg.summary$adjr2)
points(13, reg.summary$adjr2[13], col = "red", cex = 2, pch = 20)
# Cp
plot(reg.summary$cp, xlab = "Number of Variables", ylab = "Cp", type = "l")
min.cp = which.min(reg.summary$cp)
points(11, reg.summary$cp[11], col = "red", cex = 2, pch = 20)
# BIC
plot(reg.summary$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")
min.bic = which.min(reg.summary$bic)
points(11, reg.summary$bic[11], col = "red", cex = 2, pch = 20)
```

# Test MSE associated with the best model of each size

```{r}
set.seed(1)
# Number of predictors
p = 15
# Model Matrix
test.mat = model.matrix(Count ~ . -ID, data = df.test, nvmax = p)
# Vector to store test MSE
val.errors = rep(NA, p)
# Looping through number of predictors
for (i in 1:p) {
  # Coefficients of models with i predictors
  coefi = coef(regfit.full, id = i)
  pred = test.mat[,names(coefi)] %*% coefi
  val.errors[i] = mean((pred - df.test$Count)^2)
}
# Minimum Value
which.min(val.errors)
# Coefficient Values
coef(regfit.full, which.min(val.errors))
# Plotting test MSE
plot(val.errors, xlab = "# of predictors", ylab = "test MSE")
```

```{r}
# Shows frequency of variable selected according to optimal model
plot(regfit.full, scale = "adjr2")
plot(regfit.full, scale = "Cp")
plot(regfit.full, scale = "bic")
```

# New Linear Regression Model After Best Subset

```{r}
# Model
lm.mod = lm(Count ~ . -ID -Snowfall, data = df.train)
summary(lm.mod)
# MSE
lm.pred = predict(lm.mod, newdata = df.test)
MSE.lm.best = mean((df.test$Count - lm.pred)^2)
MSE.lm.best
```

# Ridge Regression

```{r}
set.seed(1)
# Cross-Validation to find lambda
y.train = df.train$Count
y.test = df.test$Count
train.mat = model.matrix(Count ~ . -ID, data = df.train)
test.mat = model.matrix(Count ~ . -ID, data = df.test)
cv.out0 = cv.glmnet(train.mat, y.train, alpha = 0)
plot(cv.out0)
# Best lambda
bestlam = cv.out0$lambda.min
bestlam
# Ridge Regression Model
ridge.mod = glmnet(train.mat, y.train, alpha = 0)
# Test Error 
ridge.pred = predict(ridge.mod, s = bestlam, newx = test.mat)
MSE.ridge = mean((y.test - ridge.pred)^2)
MSE.ridge
```

# Lasso Regression

```{r}
set.seed(1)
# Cross-Validation to find lambda
cv.out1 = cv.glmnet(train.mat, y.train, alpha = 1)
plot(cv.out1)
# Best lambda
bestlam = cv.out1$lambda.min
bestlam
# Lasso Regression Model
lasso.mod = glmnet(train.mat, y.train, alpha = 1)
# Test Error 
lasso.pred = predict(lasso.mod, s = bestlam, newx = test.mat)
MSE.lasso = mean((y.test - lasso.pred)^2)
MSE.lasso
# Number of non-zero coefficient estimates
lasso.coef = predict(lasso.mod, s = bestlam, type = "coefficients")[1:17,]
lasso.coef[lasso.coef != 0]
```

# Principal Components Regression

```{r}
set.seed(1)
# Cross-Validation to find M
pcr.fit = pcr(Count ~ . -ID -Snowfall, data = df.train, scale = TRUE, validation = "CV")
validationplot(pcr.fit, val.type = "MSEP")
summary(pcr.fit)
# Test MSE
pcr.pred = predict(pcr.fit, df.test, ncomp = 13)
MSE.pcr = mean((y.test - pcr.pred)^2)
MSE.pcr
```

# Partial Least Squares

```{r}
set.seed(1)
# Cross-Validation to find M
pls.fit = plsr(Count ~ . -ID -Snowfall, data = df.train, scale = TRUE, validation = "CV")
validationplot(pls.fit, val.type = "MSEP")
summary(pls.fit)
# Test MSE
pls.pred = predict(pls.fit, df.test, ncomp = 8)
MSE.pls = mean((y.test - pls.pred)^2)
MSE.pls
```

# Regression Spline

```{r warning=FALSE}
# Including all predictors with splines with knot = 3 on numerical predictors
spline.fit = lm(Count ~ Date + bs(Hour, knots=3) + bs(Temperature, knots=3) + bs(Humidity, knots=3) + bs(Wind, knots=3) + bs(Visibility, knots=3) + bs(Dew, knots=3) + bs(Solar, knots=3) + bs(Rainfall, knots=3) + Seasons + Holiday + Functioning, data = df.train)
summary(spline.fit)
# Plotting resulting fit
preds = predict(spline.fit, newdata = df.test)
MSE.spline = mean((df.test$Count - preds)^2)
MSE.spline
```

# GAM

```{r warning=FALSE}
set.seed(1)
# Loop to perform CV to select best DOF
delta = rep(NA, 5)
for (i in 1:5) {
  gam = gam(Count ~ Date + s(Hour, df=i) + s(Temperature, df=i) + s(Humidity, df=i) + s(Wind, df=i) + s(Visibility, df=i) + s(Dew, df=i) + s(Solar, df=i) + s(Rainfall, df=i) + s(Snowfall, df=i) + Seasons + Holiday + Functioning, data = df.train)
  delta[i] = cv.glm(df.train, gam, K = 10)$delta[1]
}
min(delta)
plot(1:5, delta)

# GAM on training data
gam = gam(Count ~ Date + s(Hour, df=5) + s(Temperature, df=5) + s(Humidity, df=5) + s(Wind, df=5) + s(Visibility, df=5) + s(Dew, df=5) + s(Solar, df=5) + s(Rainfall, df=5) + s(Snowfall, df=5) + Seasons + Holiday + Functioning, data = df.train)
summary(gam)
#par(mfrow = c(2, 3))
#plot(gam, se = T, col = "blue")
# MSE
preds = predict(gam, df.test)
MSE.GAM = mean((df.test$Count - preds)^2)
MSE.GAM
```

# Regression Tree

```{r warning=FALSE}
set.seed(1)
tree = tree(Count ~ . -ID -Snowfall, data = df.train)
summary(tree)
# Plotting Tree
par(mfrow = c(1,1))
plot(tree)
text(tree, pretty = 0)
# MSE
tree.pred = predict(tree, df.test)
MSE.regtree = mean((df.test$Count - tree.pred)^2)
MSE.regtree
# Cross-validation
cv = cv.tree(tree, FUN = prune.tree)
par(mfrow = c(1,2))
plot(cv$size, cv$dev, type = "b")
plot(cv$k, cv$dev, type = "b")
# Pruning
prune = prune.tree(tree, best = 11)
par(mfrow = c(1,1))
plot(prune)
text(prune, pretty = 0)
# Pruned MSE
prune.pred = predict(prune, df.test)
prune.MSE = mean((df.test$Count - prune.pred)^2)
prune.MSE
```

# Random Forest

```{r}
set.seed(1)
rf = randomForest(Count ~ . -ID -Snowfall, data = df.train, importance = TRUE)
rf.pred = predict(rf, newdata = df.test)
MSE.rf = mean((df.test$Count - rf.pred)^2)
MSE.rf
# Important Variables
importance(rf)
varImpPlot(rf)
```

# Bagging Approach

```{r}
set.seed(1)
bag = randomForest(Count ~ . -ID -Snowfall, data = df.train, mtry = 12, importance = TRUE)
bag.pred = predict(bag, newdata = df.test)
MSE.bag = mean((df.test$Count - bag.pred)^2)
MSE.bag
# Important Variables
importance(bag)
varImpPlot(bag)
```

# Plotting the MSE values of different models

```{r}
# Organizing data frame
mse.data = t(data.frame(MSE.lm0, MSE.lm.int, MSE.lm.best, MSE.ridge, MSE.lasso, MSE.pcr, MSE.pls, MSE.spline, MSE.GAM, MSE.regtree, MSE.rf, MSE.bag))
colnames(mse.data) = "Test MSE"
Methods = c("Original Linear Model", "Linear Model w/ Interaction Terms", "Linear Model with Best Subsets", "Ridge", "Lasso", "PCR", "PLS", "Spline", "GAM", "Tree", "Random Forest", "Bagging (rf)")
mse.data = as.data.frame(mse.data)
mse.data$Methods = Methods
# Plot
ggplot(mse.data, aes(x = reorder(Methods, -`Test MSE`), y = `Test MSE`)) + geom_point() + theme(axis.text.x = element_text(angle = 90))
```

# Predictions

```{r}
# Reading data set used for predictions
pred.data = read.table("test.csv", header = T, sep = ",")
pred.data$Count = NA

# Converting data types of certain columns
pred.data$Date = as.Date(pred.data$Date, "%d/%m/%Y")
pred.data$Seasons = as.factor(pred.data$Seasons)
pred.data$Holiday = as.factor(pred.data$Holiday)
pred.data$Functioning = as.factor(pred.data$Functioning)

# Transformations
# Wind: Right Skewed
pred.data$Wind = sqrt(pred.data$Wind)
# Visibility: Left Skewed
pred.data$Visibility = (pred.data$Visibility)^2
# Solar: Right Skewed
pred.data$Solar = sqrt(pred.data$Solar)
# Rainfall: Right Skewed
pred.data$Rainfall = sqrt(pred.data$Rainfall)
# Snowfall: Right Skewed
pred.data$Snowfall = sqrt(pred.data$Snowfall)

# Predictions
bag.pred = predict(bag, newdata = pred.data)
pred.data$Count = bag.pred

# Transforming back to original 
# Count
pred.data$Count = (pred.data$Count)^2
# Wind
pred.data$Wind = (pred.data$Wind)^2
# Visibility
pred.data$Visibility = sqrt(pred.data$Visibility)
# Solar
pred.data$Solar = (pred.data$Solar)^2
# Rainfall
pred.data$Rainfall = (pred.data$Rainfall)^2
# Snowfall
pred.data$Snowfall = (pred.data$Snowfall)^2
```

```{r}
# Formatting Data Set
testing_predictions_4288421 = pred.data[,-c(1:13)]
student_id = rep(4288421, 2208)
testing_predictions_4288421$student_id = student_id
```

```{r include=FALSE}
# Convert to CSV File
write.csv(testing_predictions_4288421, file = "testing_predictions_4288421.csv")
```
